---
title: "How to create a complex cohort"
output: rmarkdown::github_document
---

# ISB-CGC Community Notebooks

Check out more notebooks at our [Community Notebooks Repository](https://github.com/isb-cgc/Community-Notebooks)!

```
Title:   How to create a complex cohort
Author:  Lauren Hagen
Created: 2019-08-12
Purpose: More complex overview of creating cohorts
URL:     https://github.com/isb-cgc/Community-Notebooks/blob/master/Notebooks/How_to_create_a_complex_cohort.Rmd
Notes:   This notebook was adapted from work by Sheila Reynolds, 'How to Create TCGA Cohorts part 3' https://github.com/isb-cgc/examples-Python/blob/master/notebooks/Creating%20TCGA%20cohorts%20--%20part%201.ipynb.
```
***

# How to create a complex cohort

This notebook will construct a cohort for a single tumor type based on data availability, while also taking into consideration annotations about the patients or samples.

As you've seen already, in order to work with BigQuery, the first thing we need to is load the libraries required for using BigQuery in R:
```{r Load Libraries}
library(bigrquery)
library(tidyverse)
```

Just so that this doesn't get buried in the code below, we are going to specify our tumor-type of interest here.  In TCGA each tumor-type is also a separate *study* within the TCGA *project*.  The studies are referred to based on the 2-4 letter tumor-type abbreviation.  A complete list of all study abbreviations, with the full study name can be found on this [page](https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations).  For this particular exercise, we will look at the "Breast invasive carcinoma" study, abbreviated BRCA:
```{r}
studyName = "TCGA-BRCA"
```

More information the the BRCA study can be found [here](https://portal.gdc.cancer.gov/projects/TCGA-BRCA). In this notebook, we are going to wind up making use of all of the available data types, so let's have a look at the entire **`TCGA_hg38_data_v0`** dataset in the query below. The tables and data sets available from ISB-CGC in BigQuery can also be explored without login with the [ISB-CGC BigQuery Table Searcher](https://isb-cgc.appspot.com/bq_meta_search/).

```{r}
project <- 'your_project_number' # Insert your project ID in the ''
theDataset <- 'TCGA_hg38_data_v0'
if (project == 'your_project_number') {
  print('Please update the project number with your Google Cloud Project')
}
```

```{r}
print("Tables:")
tables<-list_tables("isb-cgc", theDataset)
tables
```

In this next code cell, we define an SQL query called **`DNU_patients`** which finds all patients in the Annotations table which have either been 'redacted' or had 'unacceptable prior treatment'. It will display the output of the query and then save the data to a tibble.

```{r}
# Create Query
DNU_patients_query <- "SELECT
                  case_barcode,
                  category AS categoryName,
                  classification AS classificationName
                FROM
                  `isb-cgc.TCGA_bioclin_v0.Annotations`
                WHERE
                  ( entity_type='Patient'
                    AND (category='History of unacceptable prior treatment related to a prior/other malignancy'
                    OR classification='Redaction' ) )
                GROUP BY
                  case_barcode,
                  categoryName,
                  classificationName
                ORDER BY
                  case_barcode"
# Run the query
DNU_patients <- bq_project_query(project, DNU_patients_query, quiet = TRUE) 
# Create a dataframe with the results from the query
DNU_patients <- bq_table_download(DNU_patients, quiet = TRUE)
# Show the dataframe
summary(DNU_patients)
```

Now we're gong to use the query defined previously in a function that builds a "clean" list of patients in the specified study, with available molecular data, and without any disqualifying annotations.

```{r}
buildCleanBarcodeList <- function( studyName, bqDataset, barcodeType, DNUList ) {
  print("in buldCleanBarcodeList ... ")
  ULists = list() # List to hold lists of barcodes
  print("  --> looping over data tables: ")
  barcodeField <- barcodeType # set the barcodeField to the barcodeType
  for (t in bqDataset) {
    currTable <- str_c("`isb-cgc.TCGA_hg38_data_v0.",t,"`",sep="") # Set the BigQuery Table
    barcodeField <- barcodeType # Reset the barcodeField for each loop
    # Check if the table is a somatic table and the barcodeType is "sample_barcode" as 
    # the somatic tables use sample_barcode_tumor instead
    if(grepl("Somatic", t) && barcodeField == "sample_barcode") {
      barcodeField <- "sample_barcode_tumor"
    }
    get_barcode_list <- ""
    # Check if the table is a Methlation table as these tables doesn't have the project_short_name column
    if (grepl("Methylation", t)) { 
      # Create the query
      get_barcode_list <- str_c("WITH a AS ( SELECT ",barcodeField, 
                                " FROM ", currTable, " ), b AS ( SELECT ",
                                barcodeField, 
                                ", project_short_name FROM `isb-cgc.TCGA_hg38_data_v0.Copy_Number_Segment_Masked`) SELECT ", 
                                barcodeField, 
                                " FROM ( SELECT a.", barcodeField, " AS ", 
                                barcodeField, 
                                ", b.project_short_name AS project_short_name FROM a JOIN b ON a.", 
                                barcodeField," = b.",barcodeField,
                                ") WHERE project_short_name='", studyName,
                                "' GROUP BY ", barcodeField, 
                                " ORDER BY ", barcodeField, sep = "")
    }
    else {
      get_barcode_list <- str_c("SELECT ", barcodeField, 
                                    " FROM ", currTable,
                                    " WHERE project_short_name='", studyName,
                                    "' GROUP BY ",barcodeField,
                                    " ORDER BY ",barcodeField,sep="")
    }
    # Run the query
    barcode_list <- bq_project_query(project, get_barcode_list, quiet = TRUE)
    # Create a dataframe with the results from the query
    barcode_list <- bq_table_download(barcode_list, quiet = TRUE)
    print(str_c("      ", t, "  --> ", nrow(barcode_list), " unique barcodes.", sep = ""))
    barcodes <- as.list(barcode_list) # make the barcodes a list
    ULists[[t]] <- barcodes # put the list of barcodes into ULists
  }
  print(str_c("  --> we have ", length(ULists), " lists to merge"))
  masterList <- list() # Create a list for the master list of barcodes
  for (a in 1:length(ULists)) {
    for (b in 1:length(ULists[[a]][[1]])) {
      if (!(ULists[[a]][[1]][[b]] %in% masterList)) {
        masterList <- c(masterList, ULists[[a]][[1]][[b]])
      }
    }
  }
  print(str_c("  --> which results in a single list with ", length(masterList), " barcodes"))
  print("  --> removing DNU barcodes")
  cleanList <- list()
  for ( aBarcode in 1:length(masterList) ) {
    if (!(masterList[[aBarcode]] %in% DNUList[[barcodeField]])) {
      cleanList <- c(cleanList, masterList[[aBarcode]])
    }
    else {
      print(str_c("      excluding this barcode ", masterList[[aBarcode]], sep = ""))
    }
  }
  print(str_c("  --> returning a clean list with ", length(cleanList), " barcodes", sep=""))
  return(cleanList)
}
```

```{r}
barcodeType <- "case_barcode"
cleanPatientList <- buildCleanBarcodeList(studyName,tables,barcodeType,DNU_patients)
```

Now we are going to repeat the same process, but at the sample barcode level.  Most patients will have provided two samples, a "primary tumor" sample, and a "normal blood" sample, but in some cases additional or different types of samples may have been provided, and sample-level annotations may exist that should result in samples being excluded from most downstream analyses.

```{r}
# there are many different types of annotations that are at the "sample" level
# in the Annotations table, and most of them seem like they should be "disqualifying"
# annotations, so for now we will just return all sample barcodes with sample-level
# annotations
DNUsamples_query <- "SELECT sample_barcode
                      FROM `isb-cgc.TCGA_bioclin_v0.Annotations`
                      WHERE (entity_type='Sample')
                      GROUP BY sample_barcode
                      ORDER BY sample_barcode"
# Run the query
DNUsamples <- bq_project_query(project, DNUsamples_query, quiet = TRUE) 
# Create a dataframe with the results from the query
DNUsamples <- bq_table_download(DNUsamples, quiet = TRUE)
# Show the dataframe
summary(DNUsamples)
```

And now we can re-use the previously defined function get a clean list of sample-level barcodes:

```{r}
barcodeType <- "sample_barcode"
cleanSampleList <- buildCleanBarcodeList (studyName, tables, barcodeType, DNUsamples)
```

Now we're going to double-check first that we keep only sample-level barcodes that correspond to patients in the "clean" list of patients, and then we'll also filter the list of patients in case there are patients with no samples remaining in the "clean" list of samples.

```{r}
finalSampleList <- list()

for (aSample in 1:length(cleanSampleList)) {
  aPatient <- str_sub(cleanSampleList[[aSample]], 1, 12)
  if (!aPatient %in% cleanPatientList) {
    print(str_c("     excluding this sample in the final pass: ", cleanSampleList[[aSample]], sep = ""))
  }
  else {
    finalSampleList <- c(finalSampleList, cleanSampleList[[aSample]])
  }
}
print(str_c(" Length of final sample list:", length(finalSampleList), sep = ""))
```

```{r}
finalPatientList = list()
for (aSample in 1:length(finalSampleList)) {
  aPatient <- str_sub(finalSampleList[[aSample]], 1, 12)
  if (!(aPatient %in% finalPatientList)) {
    finalPatientList <- c(finalPatientList, aPatient)
  }
}
print(str_c(" Lenth of final patient list: ", length(finalPatientList), sep = ""))

for (aPatient in 1:length(cleanPatientList)) {
  if (!finalPatientList[[aPatient]] %in% finalPatientList) {
    print(str_c("     --> patient removed in final pass: ", cleanPatientList[[aPatient]], sep=""))
  }
}
```

We're also interested in knowing what *types* of samples we have.  The codes for the major types of samples are:
- **01** : primary solid tumor
- **02** : recurrent solid tumor
- **03** : primary blood derived cancer
- **06** : metastatic
- **10** : blood derived normal
- **11** : solid tissue normal

and a complete list of all sample type codes and their definitions can be found [here](https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/sample-type-codes).

```{r}
finalSample <- tibble(sampleBarcodes = finalSampleList)

finalSample %>%
  mutate(type = str_match(sampleBarcodes, "-\\d{2}\\D")) %>%
  mutate(type = str_sub(type, 2, 3)) %>%
  group_by(type) %>%
  summarize(count=n())
```

Now we are going to create a tibble with all of the sample barcodes and the associated patient (participant) barcodes so that we can write this to a BigQuery "cohort" table.

```{r}
# Create the tibble
final <- finalSample %>%
  mutate(patientBarcodes = str_sub(sampleBarcodes, 1, 12))

# Check the patientBarcodes
final %>%
  group_by(patientBarcodes) %>%
  summarise(count=n()) %>%
  arrange(desc(count))
```

```{r}
# Check the number of unique patientBarcodes  
length(unique(final$patientBarcodes))
```

As a next step you may want to consider is to put the data into your GCP. An example of how to move files in and out of GCP and BigQuery can be found [here](https://github.com/isb-cgc/Community-Notebooks/tree/master/Notebooks) along with other tutorial notebooks.
