---
title: "How to quickly compare cohorts"
output: rmarkdown::github_document
---

# ISB-CGC Community Notebooks

Check out more notebooks at our [Community Notebooks Repository](https://github.com/isb-cgc/Community-Notebooks)!

```
Title:   How to quickly compare cohorts
Author:  Lauren Hagen
Created: 2020-04-09
URL:     https://github.com/isb-cgc/Community-Notebooks/blob/master/Notebooks/How_to_quickly_compare_cohorts.Rmd
Purpose: Comparing cohorts with survival curves and histograms with data from BigQuery or the WebApp.
Notes: 
```

# Overview

In this notebook, we will compare two cohorts with survival curves and feature comparisons. We will be using data from the [Lung Adenocarcinoma](https://portal.gdc.cancer.gov/projects/TCGA-LUAD) (LUAD) and [Lung Squamous Cell Carcinoma](https://portal.gdc.cancer.gov/projects/TCGA-LUSC) (LUSC) projects from the TCGA program.

This notebook can handle multiple or different cohorts from the TCGA program. The 'Load Cohort' code block can be updated to add or change the cohorts.

# Authorization, install or load packages, and create client

Before we get started, we will need to load the BigQuery module, authenticate ourselves, create a client variable, and import and load necessary libraries. The package `tidyverse` has dplyr, stringr, ggplot2, among other useful packages as part of it's package, so we are going to save time and just install and load the `tidyverse` package. For more information, see the `tidyverse` [site](https://www.tidyverse.org/).

```{r include=FALSE}
#Function to check whether package is installed
#https://stackoverflow.com/questions/9341635/check-for-installed-packages-before-running-install-packages
is.installed <- function(mypkg){
    is.element(mypkg, installed.packages()[,1])
} 
  
if (!is.installed("bigrquery")){
    install.packages("bigrquery")
}

if (!is.installed("dplyr")){
    install.packages("dplyr")
}

if (!is.installed("ggplot2")){
    install.packages("ggplot2")
}

if (!is.installed("stringr")){
    install.packages("stringr")
}

if (!is.installed("survival")){
    install.packages("survival")
}


if (!is.installed("ggfortify")){
    install.packages("ggfortify")
}
```


```{r}
library(bigrquery)
library(dplyr)
library(ggplot2)
library(stringr)
library(survival)
library(ggfortify)
project <- 'your_project_number' # Insert your project ID in the ''
if (project == 'your_project_number') {
  print('Please update the project number with your Google Cloud Project')
}
```

# Load Cohort

First we will need to load our cohort into a data frame. The cohort can either be made using the WebApp or a SQL query to the TCGA clinical tables. A cohort is a list of case barcodes from the TCGA program. The first code block can be updated to include 2 or more cohorts. We are going to use the sample queries filled in below as an example.

```{r}
# Fill in how the cohort will be created
# SQL or File
cohort_type <- "SQL"

# Load a list of case_barcodes from our WebApp

file1 = 'your_first_file_location_here' # update with your file location, if using one
# Or insert a SQL query
query1 = "
  SELECT
    case_barcode
  FROM
   `isb-cgc.TCGA_bioclin_v0.Clinical`
  WHERE
    disease_code = 'LUAD'
  "
file2 = 'your_second_file_location_here' # update with your file location, if using one
# Or insert a SQL query
query2 = "
  SELECT
    case_barcode
  FROM
   `isb-cgc.TCGA_bioclin_v0.Clinical`
  WHERE
    disease_code = 'LUSC'
  "

# Update with files or queries used
cohorts <- data.frame(Cohort_1=query1, Cohort_2=query2)
```

```{r}
# Function to create a combined labeled list of cohorts for comparison
load_cohort <- function(cohorts, cohort_type){
  final_list <- data.frame("case_barcodes" = NULL, "Cohort" = NULL)
  for (cohort in 1:ncol(cohorts)){
    if (cohort_type == "SQL") {
      cohort_request <- bq_project_query(project, cohorts[,cohort])
      cohort_table <- bq_table_download(cohort_request, quiet=TRUE)
    }
    else {
      cohort_table <- read.csv(cohorts[,cohort], skip = 1)
      cohort_table <- cohort_table$Case.Barcode
    }
    cohort_name <- names(cohorts[cohort])
    cohort_df <- data.frame("case_barcodes" = cohort_table, "cohort" = cohort_name)
    final_list <- rbind(final_list, cohort_df)
  }
  return(final_list)
}
  
```

```{r}
combined_list <- load_cohort(cohorts, cohort_type)
```

# Survival Curve
First, we will compare groups with a survival curve. We will be using a Kaplan Meier Curve from the [`survival` package](https://cran.r-project.org/web/packages/survival/index.html) and [`ggfortify` package](https://cran.r-project.org/web/packages/ggfortify/index.html).

```{r}
# Pull the vital status, days to death, and grouping column
# into a data frame
sc_query = str_c("
  SELECT
    case_barcode,
    vital_status,
    days_to_death
  FROM
    `isb-cgc.TCGA_bioclin_v0.Clinical`
  WHERE
    case_barcode IN ('", str_c(combined_list$case_barcode, collapse = "', '"),"') AND
    vital_status IS NOT NULL")
sc_query_request <- bq_project_query(project, sc_query)
survival_curve <- bq_table_download(sc_query_request)
survival_curve <- left_join(survival_curve, combined_list, key = "case_barcode")

# Fill in NAs in days_to_death with the max from the days to death
survival_curve$days_to_death[is.na(survival_curve$days_to_death)] <- max(survival_curve$days_to_death, na.rm = TRUE)
# Convert the vital status to numbers
survival_curve$vital_status <- ifelse(survival_curve$vital_status=='Alive', 0, 1)
```

```{r}
fit <- survfit(Surv(days_to_death, vital_status) ~ cohort, data = survival_curve)
autoplot(fit) +
  labs(title = "Survival Curve",
       y = "Percent Survival", 
       x = "Days") +
  theme(legend.title=element_blank())
```
## Survival Curve Summary

We now have our two cohorts plotted with a survival curve. We can see that the percent survival falls faster for Cohort 2 (LUSC) than Cohort 1 (LUAD). Below is a section on compare the case counts for different features within the data.

Listed below are some useful resources on Survival Curves:
- [Kaplan-Meier Estimator Wikipedia](https://en.wikipedia.org/wiki/Kaplan%E2%80%93Meier_estimator)
- [Kaplan Meier curves: an introduction](https://towardsdatascience.com/kaplan-meier-curves-c5768e349479)
- [Nature: A practical guide to understanding Kaplan-Meier curves by Rich J, Neely J, Paniello R, Voelker C, Nussenbaum B, Wang E](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3932959/)

# Feature Comparisons

We may also like to view the distribution of the patient's ages, gender, and vital status between the two data sets. We can do this by creating bar charts for each feature.

First, we will get the data from BigQuery for the desired features.

```{r}
# Create a query to retrieve the cohort features
hist_query = str_c("
  SELECT
    case_barcode,
    gender,
    vital_status,
    age_at_diagnosis,
    ethnicity,
    pathologic_stage
  FROM
    `isb-cgc.TCGA_bioclin_v0.Clinical`
  WHERE
    case_barcode IN ('", str_c(combined_list$case_barcode, collapse = "', '"),"')
  ")

hist_query_request <- bq_project_query(project, hist_query)
compare_download <- bq_table_download(hist_query_request)
compare <- left_join(compare_download, combined_list, key = "case_barcode")
```

Next we will want to view if there are any missing values in our cohorts.

```{r}
# View the number of missing records in the data set
colSums(is.na(compare))
```

Then, we will plot the feature on bar charts.

```{r}
# Create Age at Diagnosis Comparison
age_graph <- ggplot(data = compare, mapping = aes(age_at_diagnosis, fill = cohort)) +
  geom_histogram(bins = 10) +
  labs(title = "Age at Diagnosis Comparison",
       x = "Age at Diagnosis",
       y = "Count of Cases")
age_graph
```

```{r}
# Create Gender Comparison graph
gender_graph <- ggplot(data = compare, mapping = aes(ethnicity, fill=cohort)) +
  geom_bar(position = "dodge") +
  labs(title = "Ethnicity Comparison",
    x = "Ethnicity",
    y = "Count of Cases")
gender_graph
```

```{r}
# Create Gender Comparison graph
gender_graph <- ggplot(data = compare, mapping = aes(gender, fill=cohort)) +
  geom_bar(position = "dodge") +
  labs(title = "Gender Comparison",
    x = "Gender",
    y = "Count of Cases")
gender_graph
```

```{r}
# Create Vital Status Comparison graph
vital_status_graph <- ggplot(data = compare, mapping = aes(vital_status, fill=cohort)) +
  geom_bar(position = "dodge") +
  labs(title = "Vital Status Comparison",
    x = "Vital Status",
    y = "Count of Cases")
vital_status_graph
```

```{r}
# Create Pathology Status Comparison graph
pathology_status_graph <- ggplot(data = compare, mapping = aes(pathology_status, fill=cohort)) +
  geom_bar(position = "dodge") +
  labs(title = "Pathology Status Comparison",
    x = "Pathology Status",
    y = "Count of Cases")
vital_status_graph
```

# That's all folk!

So, that's it for a quick cohort comparison. If you need help or have a comment, reach out to us at feedback@isb-cgc.org

This notebook was inspired by the [ICGC Cohort Comparison Analysis tool](https://dcc.icgc.org/analysis)
